---
title: "Bikeshare_SpaceTime_Prediction"
subtitle: "CPLN 5080 - Spring 2026"
author: "Alejandro Duque"
date: today
format:
  html:
    code-fold: show
    code-tools: true
    toc: true
    toc-depth: 3
    toc-location: left
    theme: cosmo
    embed-resources: true
editor: visual
execute:
  warning: false
  message: false
---

This report analyzes Q3 2024 Indego bike share ridership in Philadelphia using 
space-time prediction models. Using trip data from July through September 2024, 
we build a series of progressively complex regression models to predict hourly 
demand at each station, evaluate spatial and temporal error patterns, and reflect 
on the equity implications of deploying such a system.

# Priming my Code

##Setting up the Code

```{r setup}
#| include: false

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(riem)  # For Philadelphia weather from ASOS stations

library(viridis) # color scales for visualization
library(knitr)
library(kableExtra)
library(patchwork)


library(here)
options(scipen = 999) # Get rid of scientific notation to make it look tidier
```

##Plot Priming

```{r themes}
#| include: false

plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

#Part 1: Importing and Prepping the Data

##Loading in the Indego Data

```{r load_indego}
#| include: false

# Read Q1 2024 data
indego <- read_csv(here("labs/lab_5/data/indego-trips-2024-q3.csv"))

```

##Examining The Data Structure

```{r explore_data}
#| message: false
#| warning: false

# How many trips?
cat("Total trips in Q3 2024:", nrow(indego), "\n")

# Date range
cat("Date range:", 
    min(mdy_hm(indego$start_time)), "to", 
    max(mdy_hm(indego$start_time)), "\n")

# How many unique stations?
cat("Unique start stations:", length(unique(indego$start_station)), "\n")

# Trip types
table(indego$trip_route_category)

# Passholder types
table(indego$passholder_type)

# Bike types
table(indego$bike_type)
```

## Creating Time Bins

```{r create-time-bins}
indego <- indego %>%
  mutate(
    # Parse datetime
    start_datetime = parse_date_time(start_time, orders = "m/d/Y H:M"),
    end_datetime   = parse_date_time(end_time,   orders = "m/d/Y H:M"),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego %>% select(start_datetime, interval60, week, dotw, hour, weekend))
```

##Trips over Time

```{r trips_over_time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips <- indego %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q3 2024",
    subtitle = "Late summer demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme

```

##Hourly Trips
```{r}
# Average trips by hour and day type
hourly_patterns <- indego %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme

```

##Top Stations

```{r top_stations}
# Most popular origin stations
top_stations <- indego %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations, 
      caption = "Top 20 Indego Stations by Trip Origins",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Philadelphia Data

```{r load_census}
#| include: false


# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2024,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

# Check the data
glimpse(philly_census)
```
## Mapping Philadelphia

```{r map_philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

## Joining Census Data to Stations

```{r join_census_to_stations}
# Create sf object for stations
stations_sf <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census <- indego %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.1) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme

```

## Dealing with missing data

```{r}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

```

##Get Weather Data

```{r}
#| echo: false   

weather_data <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2024-07-01",
  date_end = "2024-09-30"
)

# Process weather data
weather_processed <- weather_data %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete <- weather_processed %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete %>% select(Temperature, Precipitation, Wind_Speed))


```
##Visualizing Weather Patterns

```{r}
#| message: false
#| warning: false
#| 
ggplot(weather_complete, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q3 2024",
    subtitle = "Summer to early Fall transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme
```

#Create Space-time panel

```{r}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panel <- indego_census %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel)

# How many unique stations?
length(unique(trips_panel$start_station))

# How many unique hours?
length(unique(trips_panel$interval60))

hist(trips_panel$Trip_Count)
```
##Create Complete Panel Structure

```{r complete-panel}
# Calculate expected panel size
n_stations <- length(unique(trips_panel$start_station))
n_hours <- length(unique(trips_panel$interval60))
expected_rows <- n_stations * n_hours

cat("Expected panel rows:", format(expected_rows, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows - nrow(trips_panel), big.mark = ","), "\n")

# Create complete panel
study_panel <- expand_grid(
  interval60 = unique(trips_panel$interval60),
  start_station = unique(trips_panel$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel)
hist(study_panel$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes <- trips_panel %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes)

study_panel <- study_panel %>%
  left_join(station_attributes, by = "start_station")

glimpse(study_panel)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel), big.mark = ","), "\n")
```

## Add Time Features
```{r add_time_features}
study_panel <- study_panel %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel)
```
## Join Weather Data
```{r join_weather}
#| echo: false   

study_panel <- study_panel %>%
  left_join(weather_complete, by = "interval60")

glimpse(study_panel)

# Check for missing values
summary(study_panel %>% select(Trip_Count, Temperature, Precipitation))
```
###Feature Engineering: Holiday and TOO hot Indicator
```{r}
study_panel <- study_panel %>%
  mutate(
#1. Holiday indicator for Q3 holidays
    is_holiday = case_when(
      month(interval60) == 7 & day(interval60) == 4 ~ 1,  # July 4th
      month(interval60) == 9 & wday(interval60) == 2 & 
        day(interval60) <= 7 ~ 1,                          # Labor Day (first Monday in Sept)
      TRUE ~ 0
    ),
    
#Perfect biking weather (65-80°F, no rain)
    perfect_weather = ifelse(
      Temperature >= 65 & Temperature <= 80 & Precipitation == 0, 1, 0
    ),
    #Too hot indicator (>88°F suppresses ridership in summer)
    too_hot = ifelse(Temperature > 88, 1, 0)
  )
```


# Creating Temporal Lag Variables

```{r create_lags}
# Sort by station and time
study_panel <- study_panel %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel <- study_panel %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24),
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete <- study_panel %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete), big.mark = ","), "\n")
```
## Visualize Lag Correlations

```{r lag_correlations}
# Sample one station to visualize
example_station <- study_panel_complete %>%
  filter(start_station == first(start_station)) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

# Temporal Train / Test Split

```{r temporal-split}
# Split by week
# Q1 has weeks 1-13 (July-Sept)
# Train on weeks 1-9 (July 1 - early Sept)
# Test on weeks 10-13 (rest of Sept)

# Which stations have trips in BOTH early and late periods?
early_stations <- study_panel_complete %>%
  filter(week < 37) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations <- study_panel_complete %>%
  filter(week >= 37) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations <- intersect(early_stations, late_stations)


# Filter panel to only common stations
study_panel_complete <- study_panel_complete %>%
  filter(start_station %in% common_stations)

# NOW create train/test split
train <- study_panel_complete %>%
  filter(week < 37)

test <- study_panel_complete %>%
  filter(week >= 37)

cat("Training observations:", format(nrow(train), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test), big.mark = ","), "\n")
cat("Training date range:", format(min(train$date), "%B %d, %Y"), "to", format(max(train$date), "%B %d, %Y"), "\n")
cat("Testing date range:", format(min(test$date), "%B %d, %Y"), "to", format(max(test$date), "%B %d, %Y"), "\n")

```
#Buidling Predictive Modeling 

## Model 1: Baseline (Time + Weather)

```{r model1}


# Create day of week factor with treatment (dummy) coding
train <- train %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))


# Now run the model
model1 <- lm(
 Trip_Count ~ as.factor(hour) + factor(dotw) + Temperature + Precipitation,
  data = train
)

summary(model1)
```

## Model 2: Adding Temporal Lags


```{r model2}

model2 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train
)

summary(model2)
```
## Model 3: Adding Demographics

```{r}
model3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train
)

summary(model3)
```
## Model 4: Add Station Fixed Effects
```{r model4}
model4 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station),
  data = train
)

# Summary too long with all station dummies, just show key metrics
cat("Model 4 R-squared:", summary(model4)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4)$adj.r.squared, "\n")
```

## Model 5: Add Rush Hour Interaction
```{r model5}
model5 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train
)

cat("Model 5 R-squared:", summary(model5)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5)$adj.r.squared, "\n")
```

## Calculating Predictions and MAE

```{r calculate_mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Visualize Model Comparison
```{r}
ggplot(mae_results, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Error Analysis

## Observed vs. Predicted Values

```{r obs_vs_pred}
#| message: false
#| warning: false

test <- test %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips",
    subtitle = "Model 2 performance by time period",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme
```

## Spatial Error Patterns

```{r spatial_errors}
# Calculate MAE by station
station_errors <- test %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors <- test %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))




# Map 1: Prediction Errors
p1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE (trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand  
p2 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.1) +
  geom_point(
    data = station_errors,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg Demand (trips/hour)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand") +
  mapTheme +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  guides(color = guide_colorbar(
    barwidth = 12,
    barheight = 1,
    title.position = "top",
    title.hjust = 0.5
  ))

# Combine
(p1 | p2)

```

There are many reasons as to why the spatial model produces errors across the city. In Center City we see high error and high demand, some possible reasons being that the model does not know when event based surges in ridership occur and the unpredictability of tourism surges also plays into the uncertainty. 

Between July and late September, university students start to move in for the start of the school year which also plays into the factor of unpredictability. 


## Temporal Error Patterns

```{r temporal_errors}
# MAE by time of day and day type
temporal_errors <- test %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We find the highest errors during PM Rush, AM Rush on the weekday, and midday rush. PM and AM rushes have high levels of error, because this is when the model sees the most demand in ridership -- yet, a rainy Tuesday morning, compared to a sunny Friday morning, you will have different ridership numbers, the model does not know this!

## Demographic Error Patterns

```{r errors_demographics}
#| message: false
#| warning: false

# Join demographic data to station errors
station_errors_demo <- station_errors %>%
  left_join(
    station_attributes %>% select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

# Create plots
p1 <- ggplot(station_errors_demo, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income", x = "Median Income", y = "MAE") +
  plotTheme

p2 <- ggplot(station_errors_demo, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3 <- ggplot(station_errors_demo, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race", x = "% White", y = "MAE") +
  plotTheme

(p1 | p2) / (p3 | plot_spacer())
```
According to the model, higher median income neighborhoods are harder to predict for, citing higher levels of MAE. This could be that these wealthier neighborhoods have busier stations, with more unpredictable patterns in ridership. Conversely, when the percentage of residents who take transit increase, the MAE decreases, probably because those neighborhoods who are likely to commute on transit are also more likely to use Indego. Commuting is Predictable!

Because wealthier, higher deman neighborhoods also tend to have higher percentages of White residents, MAE in these areas were seen to be high. This makes sense, consistently neighborhoods that are wealthier, have high ridership and more erratic demand patterns that can be sometimes more based on events. Whereas low demand areas are just as hard to predict for. Sparse stations, and lower ridership, make for irregular commuting patterns and the model has less trips to train on.


#Feature Engineering & Model Improvement

## Holiday and Good Weather Biking Model

```{r}
model6 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    as.factor(start_station) +
    rush_hour * weekend +
    is_holiday + perfect_weather + too_hot,
  data = train
)

cat("Model 6 R-squared:", summary(model6)$r.squared, "\n")
cat("Model 6 Adj R-squared:", summary(model6)$adj.r.squared, "\n")
```

##Calculating Model 6 MAE and comparing between 

```{r new_calculate_mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test <- test %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test$dotw_simple) <- contr.treatment(7)

test <- test %>%
  mutate(
    pred1 = predict(model1, newdata = test),
    pred2 = predict(model2, newdata = test),
    pred3 = predict(model3, newdata = test),
    pred4 = predict(model4, newdata = test),
    pred5 = predict(model5, newdata = test),
    pred6 = predict(model6, newdata = test)
  )

# Calculate MAE for each model
mae_results <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction",
    "6. + Holidays and Good Weather"
  ),
  MAE = c(
    mean(abs(test$Trip_Count - test$pred1), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred2), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred3), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred4), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred5), na.rm = TRUE),
    mean(abs(test$Trip_Count - test$pred6), na.rm = TRUE)
  )
)

kable(mae_results, 
      digits = 2,
      caption = "Mean Absolute Error by Model (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

Two key elements were missing from the previous models, a holiday indicator and a good biking weather indicator. Model 6 aims to include some of these elements in the Q3 2024 Indego data. During Q3, July 4th and Labor Day are two holidays that would break the ridership patterns of typical Philadelphia days. While the perfect weather indicator found hours when temperatures fell between 65 and 85 degrees. 

Although adding these features are theoretically backed up, they did not improve the model's performance to a substantial degree. In fact, no change in MAE was seen from model 5 to model 6. This is likely due to the fact that the holiday indicator is limited to two days across the whole training data period and even though the perfect weather indicator specified hours, weather was already used in the previous regressions.


#Critical Reflection


A MAE of 0.70 trips per station-hour sounds impressive on paper, but whether it is "good enough" for Indego to actually use depends heavily on context. For a low demand station averaging 1-2 trips per hour, being off by 0.70 trips is a significant error, that's essentially not knowing whether a bike will be needed at all. For a high demand Center City station averaging 5-6 trips per hour, 0.70 is much more acceptable. So the model's usefulness really depends on which stations Indego is trying to optimize rebalancing for. Prediction errors become most damaging during peak hours — AM and PM rush on weekdays — when a wrong prediction means a dock is either empty or full at exactly the wrong time. I would recommend deploying this system only as a decision support tool rather than a fully automated rebalancing system, and only for high-demand stations where the model has enough training signal to be reliable.

On the equity side, the spatial error analysis raises some real concerns. Prediction errors are not distributed evenly across Philadelphia — stations in lower-income neighborhoods tend to have less predictable demand patterns, which means the rebalancing algorithm informed by this model will consistently deprioritize those areas. This could worsen existing disparities in bike access, where communities that already have fewer transportation options end up with less reliable bike share service. To address this, I would recommend setting minimum service guarantees for lower-income stations that operate independently of model predictions, and regularly auditing errors by census tract to catch any systematic bias before it compounds over time.

The model has a few meaningful limitations worth acknowledging. The biggest is that it cannot capture event-driven demand, like a Phillies game, a festival on the parkway, or a large convention will all cause spikes that look like noise to this model. It also assumes that the relationship between weather, time, and demand stays stable, which does not hold well across the summer-to-fall transition in Q3 where demand patterns shift fairly quickly in September.

-------------------------------------------------
